Comparing ChatGPT-4 to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-23 manuscript/manuscript.tex
Add overview of models, now working on healsl study

\abstract{\textbf{Background:} Verbal Autopsies (VAs) collect data on deaths and their causes outside of traditional hospital settings to provide more representative counts and Causes of Death (CODs) for reducing premature mortality. Current computer models for COD assignment in VAs perform similar to physicians at the population level, but poorly at the individual level, due to a focus on structured questionnaire data and neglecting free text from the open narratives. Recently, a Generative Pre-trained Transformer (GPT) model called ChatGPT has demonstrated human-level performance on professional and academic exams using free text input. ChatGPT shows promise in mimicking physician behavior for assigning CODs, but to the best of our knowledge, has yet to be tested for assigning CODs using open narratives from VAs.

\textbf{Methods:} 6942 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/ChatGPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC among models were evaluated for all records and by COD, age group, and age ranges.

@@

In 2019, 41 million people die prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \cite{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable, but require adequate resource allocation, guided by evidence, to implement effective interventions and policies that target populations at risk \cite{benzigerGlobalBurdenDisease2016}. Thus, reliable counts and diagnoses of deaths provide decision makers with evidence to save lives and reduce premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries do not have data on deaths or have registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of deaths occur at home \cite{adairWhoDiesWhere2021}.

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. Although VA surveys have been an effective alternative to collect mortality data at scale (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), COD assignment has been criticized to be expensive and difficult to reproduce due to reliance on physician assignment \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. As an alternative to physician assignment, computer models, such as InterVA \cite{byassIntegratedApproachProcessing2019} and InSilicoVA \cite{mccormickProbabilisticCauseofDeathAssignment2016}, have recently been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \cite{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021}. These computer models often utilized data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \cite{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023a}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare the performances of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

@@

Initially, 11,920 records were collected from dual-coded Electronic Verbal Autopsy (EVA), where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement, codes from two physicians per record were compared in the EVA system to determine if they agreed using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes are not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes, and assigns a final code. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated), resulting in 6942 physician agreed records. Since computer models were compared to physicians, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}.

@@

Four computer models were used to predict COD for each of the 6942 physician agreed records: ChatGPT-3.5 \cite{brownLanguageModelsAre2020}, ChatGPT-4 \cite{openaiGPT4TechnicalReport2023a}, InterVA-5 \cite{byassIntegratedApproachProcessing2019}, and InSilicoVA \cite{mccormickProbabilisticCauseofDeathAssignment2016}. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs.

@@

For ChatGPT-3.5 and ChatGPT-4, text prompts were generated for each record as input data with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

\subsubsection{Metrics by Age and Cause of Death}\label{methods-eval-agecod}

The CSMF accuracy and PCCC metrics were calculated for each model by age group, and further into age ranges and cause of death for each age group to examine model performance for particular ages and CODs. Metrics were calculated overall for three age groups: adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days). For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. Metrics were also calculated by CODs defined by CGHR-10 codes, which include 19, 10, and 7 CODs for adult, child, and neonatal records.

>>>>

\abstract{\textbf{Background:} Verbal Autopsies (VAs) collect data on deaths and their causes outside of traditional hospital settings to provide more representative counts and Causes of Death (CODs) for reducing premature mortality. Current computer models for COD assignment in VAs perform similar to physicians at the population level, but poorly at the individual level, due to a focus on structured questionnaire data and neglecting free text from the open narratives. Recently, a Generative Pre-trained Transformer (GPT) model called ChatGPT-4 has demonstrated human-level performance on professional and academic exams using free text input. ChatGPT-4 shows promise in mimicking physician behavior for assigning CODs, but to the best of our knowledge, has yet to be tested for assigning CODs using open narratives from VAs.

\textbf{Methods:} 6942 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/ChatGPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC to physicians among models were evaluated for all records and by COD, age group, and age ranges.

@@

In 2019, 41 million people die prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \cite{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable, but require adequate resource allocation, guided by evidence, to implement effective interventions and policies that target populations at risk \cite{benzigerGlobalBurdenDisease2016}. Thus, reliable counts and diagnoses of deaths enable decision makers to identify populations at risk to save lives and reduce premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries do not have data on deaths or have registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of deaths occur at home \cite{adairWhoDiesWhere2021}.

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. Although VA surveys have been an effective alternative to collect mortality data at scale (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), COD assignment has been criticized to be expensive and difficult to reproduce due to reliance on physician assignment \cite{mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. As an alternative to physician assignment, computer models, such as InterVA \cite{byassIntegratedApproachProcessing2019} and InSilicoVA \cite{mccormickProbabilisticCauseofDeathAssignment2016b}, have recently been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \cite{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}. These computer models often utilized data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \cite{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

\subsection{Healthy Sierra Leone (HEAL-SL) Study Dataset}\label{methods-healsl}

The HEAL-SL study aims to \cite{carshon-marshChildMaternalAdult2022}.

@@

The initial 11,920 records were collected from dual-coded Electronic Verbal Autopsy (EVA), where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated), resulting in 6942 physician agreed records. Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}.

@@

Four computer models were used to predict COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation.

\subsubsection{Overview of ChatGPT-3.5/4, InterVA-5, and InSilicoVA}\label{methods-models-overview}

ChatGPT-3.5 \cite{brownLanguageModelsAre2020} and ChatGPT-4 \cite{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in ChatGPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. ChatGPT-3.5 and ChatGPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} standard statistical models for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, ChatGPT-3.5 and ChatGPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

@@

For ChatGPT-3.5 and ChatGPT-4, text prompts were generated for each record as input to instruct the models to assign CODs based on the open narratives. These prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

\subsubsection{Metrics Overall and by Age and Cause of Death}\label{methods-eval-agecod}

The CSMF accuracy and PCCC metrics were calculated and compared for each model overall and by age group, followed by age ranges and cause of death for each age group as model performance can vary across ages and specific causes \cite{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}. Metrics were calculated overall for three age groups: adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days). For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. Metrics were also calculated by CODs defined by CGHR-10 codes, which include 19, 10, and 7 CODs for adult, child, and neonatal records respectively.
